---
title: "UsingStackImpute"
author: "Created by Dr. Lauren J Beesley. Contact: lbeesley@umich.edu"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{UsingStackImpute}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



In this vignette, we provide a brief introduction to using the R package *StackImpute*. The purpose of this package is to provide resources for performing data analysis in the presence of missing data through analysis of a stacked version of the data. We will not include technical details about this estimation approach and instead focus on implementation. For additional details about the estimation algorithm, we refer the reader to Beesley et al. (2020) in *UNDER REVIEW*.



## Example analysis using standard imputation stacking 

Many researchers have noted that we can do a "good" job estimating regression model parameters by analyzing stack from a stacked version of the imputed data. However, estimation of standard errors has proven to be a challenging problem. We provide an estimation strategy and corresponding software for routine estimation as follows. 

```{r, echo = TRUE, eval = FALSE,  fig.width = 7, fig.height= 4}
library(dplyr)
### Simulate Data
Nobs = 2000
DAT = MASS::mvrnorm(n = Nobs, mu = c(0,0,0), Sigma = rbind(c(1, 0.18, 0.42), c(0.18, 0.09, 0.12),c(0.42, 0.12, 0.49 )))
Y = DAT[,1]
B = DAT[,2]
X = DAT[,3]
S = sample(x=c(0,1), size = Nobs, prob = c(0.5,0.5), replace = TRUE)
complete_cases = data.frame(Y, X, B, S)[S == 1,] #complete case subjects only
observed_data = data.frame(Y, X, B, S) #data with missingness in B
observed_data[S==0,'B'] = NA

### Step 1: Impute B|X,Y
imputes = mice::mice(observed_data, m=50, method="norm", printFlag=F, maxit = 1)
pred = imputes$predictorMatrix 
pred[pred != 0] = 0
pred["B","X"] = 1
pred["B","Y"] = 1
imputes = mice::mice(observed_data, m=50, predictorMatrix=pred, method="norm", printFlag=F)

### Step 2: Stack imputed datasets	
stack = mice::complete(imputes, action="long", include = FALSE)

### Step 3: Obtain weights
stack$wt = 1
stack = as.data.frame(stack %>% group_by(.id) %>% mutate(wt = wt / sum(wt)))

### Step 4: Estimation
fit = glm(Y ~X + B, data=stack, family=gaussian(), weights = stack$wt)
Info = StackImpute::Louis_Information(fit, stack, M = 50, IMPUTED = unique(stack$.id[stack$S==0]))
VARIANCE = diag(solve(Info))
```



## Example analysis using modified imputation stacking allowing for compatibility of the imputation and analysis models

In Beesley et al. (2020) in *UNDER REVIEW*, we propose a modification to this data analysis pipeline that involves imputing missing covariates from distributions that do not condition on the outcome of interest and then performing a weighted analysis on the stacked data. In this case, weights are defined proportion to the distribution of the outcome given covariates in the complete case data. We can implement this modified analysis pipeline as follows: 


```{r, echo = TRUE, eval = FALSE,  fig.width = 7, fig.height= 4}
### Step 1: Impute B|X
imputes = mice::mice(observed_data, m=50, method="norm", printFlag=F, maxit = 1)
pred = imputes$predictorMatrix 
pred[pred != 0] = 0
pred["B","X"] = 1
imputes = mice::mice(observed_data, m=50, predictorMatrix=pred, method="norm", printFlag=F)

### Step 2: Stack imputed datasets	
stack = mice::complete(imputes, action="long", include = FALSE)

### Step 3: Obtain weights
fit_cc = glm(Y ~ X + B, family='gaussian', data= complete_cases)
stack$wt = dnorm(stack$Y,mean = predict(fit_cc, newdata = stack), sd = sqrt(summary(fit_cc)$dispersion))
stack = as.data.frame(stack %>% group_by(.id) %>% mutate(wt = wt / sum(wt)))

### Step 4: Estimation
fit = glm(Y ~X + B, data=stack, family=gaussian(), weights = stack$wt)
Info = StackImpute::Louis_Information(fit, stack, M = 50, IMPUTED = unique(stack$.id[stack$S==0]))
VARIANCE = diag(solve(Info))
```



## A note on stack structure

We can equivalently performed our proposed data analysis using a stacked dataset with M rows for each subject (with M being the number of multiple imputations) or using a stacked dataset in which subjects with fully-observed data only appear once. The advantage of this latter approach is that estimation may be faster and the stacked dataset may be much smaller. We can perform the above analysis using the shorter stack obtained below 
```{r, echo = TRUE, eval = FALSE,  fig.width = 7, fig.height= 4}
### Step 2: Stack imputed datasets	
stack = mice::complete(imputes, action="long", include = FALSE)
cc = unique(stack$.id[stack$S == 1])
stack_short = rbind(stack[stack$S==0,], stack[stack$S==1 & !duplicated(stack$.id),])
```



